---
title: "Getting started with torch"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(torch)
library(luz)
library(dplyr)
knitr::opts_chunk$set(echo = FALSE)
```

## Welcome to `torch`! (And `luz`!)

### `torch`

[`torch`](http:/torch.mlverse.org) is an *R-native* framework for fast array computation with automatic differentiation and rich neural-network functionality.

TBD pic

### `luz`

What `keras` is to `tensorflow` -- a high-level API streamlining and instrumenting the training process -- `luz` is to `torch`. While everything can be accomplished with `torch` alone, `luz`

-   provides a *keras*-like, declarative interface to the training process

```{=html}
<!-- -->
```
-   removes boilerplate code

-   keeps track of *metrics* commonly used in data science, and allows you to define your own

-   provides a set of ready-to-use *callbacks* to control training, save model state, and more

-   offers a convenient interface for defining your own callbacks

### Ecosystem

-   [torch](https:/github.com/mlverse/torch)

-   [luz](https:/github.com/mlverse/luz)

```{=html}
<!-- -->
```
-   [torchvision](https:/github.com/mlverse/torchvision)

-   [torchdatasets](https:/github.com/mlverse/torchdatasets)

-   [tabnet](https:/github.com/mlverse/tabnet)

-   ... and more!

### Our goals for today

1.  Understand and use `torch` tensors and neural network modules; understand and apply automatic differentiation.

2.  Use `luz` to train neural networks in a declarative way.

3.  Get started with time-series forecasting in `torch`.

## `torch` tensors, modules, and autograd

### Tensors

#### Creating tensors

##### Way 1: From R values

Tensors can be created directly from R values using `torch_tensor()`. Optionally, we can define tensor *attributes*, including the data type, the device it lives on, and more.

Here we are creating one-dimensional tensors (vectors):

```{r, eval=FALSE, echo=TRUE}
torch_tensor(1)
torch_tensor(1, dtype = torch_int())
torch_tensor(1, device = "cuda")

torch_tensor(c(1, 2, 3)) # float tensor
```

Two-dimensional tensors can be created from R matrices.

```{r, eval=FALSE, echo=TRUE}
torch_tensor(matrix(1:9, ncol = 3)) # integer tensor
torch_tensor(matrix(1:9, ncol = 3))$to(dtype = torch_float()) # cast to float

torch_tensor(matrix(1:9, ncol = 3, byrow = TRUE))
```

Higher-dimensional tensors can be created from R arrays, but normally it is easier to use bulk creation functions.

##### Way 2: Bulk creation functions.

Multi-dimensional tensors following some defineable pattern are created passing in the desired dimensionality. A few examples (more exist):

```{r, eval=FALSE, echo=TRUE}
torch_zeros(c(3, 3))
torch_rand(c(3, 3))

```

Another often-used type of function is used specifying the desired range:

```{r, eval=FALSE, echo=TRUE}
torch_arange(1, 9)
torch_logspace(start = 0.1, end = 1.0, steps = 5)
```

#### Converting back to R

Tensors are converted back to R using `as.numeric()`, `as.matrix()`, or `as.array()`:

```{r, eval=FALSE, echo=TRUE}
torch_tensor(2) %>% as.numeric()

torch_ones(c(2, 2)) %>% as.matrix() 

torch_ones(c(2, 2, 2)) %>% as.array() 
```

#### Operations on tensors

A great number of operations can be performed on tensors. In general, there is a pair of *function* (not associated to any object) and corresponding *method* ("belonging" to a tensor instance) that do the same thing:

```{r, eval=FALSE, echo=TRUE}
t1 <- torch_tensor(c(1, 2, 3))
t2 <- torch_tensor(c(1, 2, 3))

torch_add(t1, t2)
t1$add(t2)
```

In both cases, the original tensors are not modified; a new object is created. Normally, you would just assign this to a new variable:

```{r, eval=FALSE, echo=TRUE}
t3 <- t1$add(t2)

t1
t3
```

In the few cases where you need to modify the original tensor, you can make use of the corresponding underscore variants:

```{r, eval=FALSE, echo=TRUE}
t1$add_(t2)
t1
```

Here are a few of the many matrix operations available. `$mul()` multiplies element-wise; `$matmul()` performs matrix multiplication; `$dot()` computes the dot product:

```{r, eval=FALSE, echo=TRUE}
t1$mul(t2)

# both work (torch has no concept of row vs. column vector)
t1$matmul(t2)
t1$t()$matmul(t2)

t1$dot(t2)
```

You can see that `torch` makes no distinction between row and column vectors. Above, `$t()` transposes the vector `t1`, but the matrix multiplication will work without.

#### Reshaping tensors

Often, you will need to reshape a tensor. Among the most common operations are `$squeeze()` and `$unsqueeze()`. The former adds removes a singleton dimension at the specified position (where singleton means the dimension is of length `1`):

```{r, eval=FALSE, echo=TRUE}
t1 <- torch_randn(c(1, 2, 3, 4))
t1

t1$squeeze(1)
```

The latter, in contrast, adds a singleton dimension:

```{r, eval=FALSE, echo=TRUE}
t1$unsqueeze(4)
```

This only works for singleton dimensions. `$view()` works for arbitrary reshaping, provided the number of elements allows for it. `t1` , above, has 24 values, which could as well be arranged as 6x4 or 1x24:

```{r, eval=FALSE, echo=TRUE}
t1$view(c(6, 4))

t1$view(24)

```

`$view()` does not actually create a separate, re-shaped instance of its input tensor; instead, it has the new variable refer to the same location in memory, and just stores some metadata that tell `torch` how the respective bytes should be read. There are cases when `$view()` cannot be used; in this case, you can always use `$reshape()` instead. In contrast to `$view()`, `$reshape()` will make a physical copy if necessary.

#### Indexing and slicing

Indexing in `torch` is 1-based, just like in R overall. And just like in R, singleton dimensions will be dropped -- unless you specify `drop = FALSE`:

```{r, eval=FALSE, echo=TRUE}
t1

t1[ , 1, , ]
t1[ , 1, , , drop = FALSE]

```

Ranges of values ("slices") can be accessed using the semicolon:

```{r, eval=FALSE, echo=TRUE}
t1[1, 1, 1:2, ]
t1[1, 1, 1:2, , drop = FALSE]
```

A shortcut that does not exist in R (where the same syntax has different semantics), index `-1` is used to refer to the last element in a dimension:

```{r, eval=FALSE, echo=TRUE}
t2 <- torch_tensor(1:17)
t2[-1] 
```

#### Broadcasting

In `torch`, tensors may be *broadcasted*. The principle is the same as when, in R, we add a scalar to every element in a vector. But it goes farther than that. We don't have the time to explain the rules in detail, but show a few examples as well as state the rules, for you to return at a later time.

Here, we "add" a matrix and a vector, resulting in the vector being added to every *row* of the matrix. This is possible only because `t2` has a singleton dimension in front.

```{r, eval=FALSE, echo=TRUE}
t1 <- torch_randn(c(3,5))
t2 <- torch_randn(c(1,5))

t1$add(t2)
```

This example looks similar, but it involves an additional operation from `torch`'s side: `t2` is first virtually expanded to size 1x5 (a singleton dimension is added in front). Then, things go like above.

```{r, eval=FALSE, echo=TRUE}
t1 <- torch_randn(c(3,5))
t2 <- torch_randn(c(5))

t1$add(t2)
```

As a final example, here we see both virtual addition of a singleton dimension (to `t1`) and the "reusability" of singleton dimensions shown in the first example. The latter idea is used twice, for `t1` as well as `t2`.

#### Appendix: Broadcasting rules

    # 1 We align array shapes, starting from the right.
      
      # Example

      # t1, shape:     8  1  6  1
      # t2, shape:        7  1  5
      

    # 2 Starting to look from the right, the sizes along aligned axes either have to match exactly,
    #   or one of them has to be equal to 1.
    #   In the latter case, the 1-dimensional tensor is broadcast to the larger one.

      # Example: this happens in the last (for t1) as well as the second-from-last dimension (for t2)

      # t1, shape:     8  1  6  5
      # t2, shape:        7  6  5


    # 3 If on the left, one of the arrays has an additional axis (or more than one),
    #   the other is virtually expanded to have a size of 1 in that place.
    #   Then, broadcasting will happen as stated in (2).

      # Example: this happens in t1’s leftmost dimension. First, there is a virtual expansion

      # t1, shape:     8  1  6  1
      # t2, shape:     1  7  1  5

      # and then, broadcasting happens:
      
      # t1, shape:     8  1  6  1
      # t2, shape:     8  7  1  5

#### Exercise: Tensors

In the following exercises, try translating the R code into equivalent operations using `torch`.

1.  Create two tensors representing a matrix and a vector, respectively:

```{r tensors1, exercise=TRUE, exercise.eval=TRUE}
# a matrix
m1 <- matrix(1:32, ncol = 8, byrow = TRUE)

# really a vector
m2 <- matrix(1:8, ncol = 1)

m1
m2
```

```{r tensors1-hint}
t1 <- torch_tensor(matrix(1:32, ncol = 8, byrow = TRUE))
t2 <- torch_tensor(1:8)

t1
t2
```

2.  Multiply the matrices, sum over all elements, and take the square root:

```{r tensors2, exercise=TRUE, exercise.eval=TRUE}
(m1 %*% m2)^2 %>% sum() %>% sqrt()
```

```{r tensors2-hint}
t1$matmul(t2)$square()$sum()$to(dtype = torch_float())$sqrt()
```

[Note how we need to cast to `float` in order to be able to call `torch_sqrt()`.]

3.  Multiply each row in `m1` by the vector `m2` (element-wise):

```{r tensors3, exercise=TRUE, exercise.eval=TRUE}
m1 * rbind(t(m2), t(m2), t(m2), t(m2))
```

```{r tensors3-hint}
t1 * t2
```

[Note how broadcasting takes care of the duplication for us. Also, note how no transposition is needed, as `torch` has no concept of row vectors vs. column vectors.]

4.  Transpose the matrix `m1`, and compute column sums. (This should yield 4 values.)

```{r tensors4, exercise=TRUE, exercise.eval=TRUE}
t(m1) %>% apply(2, sum)
```

```{r tensors4-hint}

t1$t()$sum(dim = 1)
```

[Note how applying the sum over dimension 1 (not 2) collapses the rows. Try to view it like this: Given an index into the dimensions, in R, we think "group by". In torch, we think "collapse".]

5.  Standardize `m1` , subtracting the mean and dividing by the standard deviation.

```{r tensors5, exercise=TRUE, exercise.eval=TRUE}
(m1 - mean(m1)) / sd(m1)
```

```{r tensors5-hint}
t1 <- t1$to(dtype = torch_float())
(t1 - t1$mean()) / t1$std()
```

Just like `torch_sum()`, `torch_mean()` and `torch_std()` need their input to be of type `float`.

### Automatic differentiation with *autograd*

#### How it works

`torch` autograd provides automatic differentiation for operations executed on tensors. For this to happen, the "source" (or "leaf", as `torch` calls it) tensor -- the one *with respect to which* we'd like derivatives computed -- needs to be created with `requires_grad = TRUE`. Let's call it `a`:

```{r, eval=FALSE, echo=TRUE}
a <- torch_tensor(matrix(1:4, ncol = 2, byrow = TRUE), dtype = torch_float(), requires_grad = TRUE)

```

In this example, `c`, the output, depends on `a` via `b`:

```{r, eval=FALSE, echo=TRUE}
b <- a$mul(2)
c <- b$sum()
```

So far, no derivatives have been computed yet. But `torch` knows what to do should we ask it to. More precisely, it knows the concrete operations it'll have to compute the derivatives for:

```{r, eval=FALSE, echo=TRUE}
c$grad_fn
b$grad_fn
```

To actually have them computed, call `$backward()` on the output tensor:

```{r, eval=FALSE, echo=TRUE}
c$backward()
```

Now the gradient of `c` with respect to `a` can be found in `a`'s `$grad` field.

```{r, eval=FALSE, echo=TRUE}
a$grad
```

When we're updating a "leaf" tensor, for example in optimization, we don't want `torch` to record that operation for later computation of derivatives. In these cases, we need to tell it to exempt the operation in question from the process:

```{r, eval=FALSE, echo=TRUE}
     
with_no_grad( {
  a$sub_(0.1 * a$grad)
})

a
```

#### Minimizing a function with *autograd*

We can use *autograd* to minimize a function. We define a parameter to hold $\mathbf{x}$. Then, in a loop, we evaluate the function at the current $\mathbf{x}$, compute the gradient, and subtract a fraction of the gradient from $\mathbf{x}$.

```{r, eval=FALSE, echo=TRUE}

# function to minimize
f <- function(x) x^2 - 7

# we start from x = 11
param <- torch_tensor(11, requires_grad = TRUE)

# learning rate: fraction of gradient to subtract
lr <- 0.1

for (i in 1:num_iterations) {
  
  # call function on current parameter value

  # compute gradient of value w.r.t. parameter

  # update parameter

}
```

In the exercise, you're asked to fill in the missing pieces.

#### Exercise: Function minimization

Fill in the lines marked "TBD". When you have the code running, experiment with the learning rate and compare the results. What is a good learning rate for this problem?

```{r autograd, exercise=TRUE, exercise.eval=TRUE}
# function to minimize
f <- function(x) x^2 - 7

# we start from x = 11
param <- torch_tensor(11, requires_grad = TRUE)

# learning rate: fraction of gradient to subtract
lr <- 0.1

for (i in 1:10) {
  
  cat("Iteration: ", i, "\n")
  
  # call function with current parameter
  value <- 777 # TBD
  cat("Value is: ", as.numeric(value), "\n")
  
  # compute gradient of value w.r.t. parameter
  # TBD
  # uncomment the followig line when ready
  # cat("Gradient is: ", as.matrix(param$grad), "\n")
  
  # update parameter
  # wrap in with_no_grad
  with_no_grad({
    # subtract a fraction of gradient from param
    # TBD
    
    # zero out on every iteration (would accumulate otherwise)
    # TBD
  })
  
  cat("After update: Param is: ", as.matrix(param), "\n\n")
  
  if (abs(-7 - as.numeric(value)) < 0.00005) break
}
```

```{r autograd-hint}
# function to minimize
f <- function(x) x^2 - 7

# we start from x = 11
param <- torch_tensor(11, requires_grad = TRUE)

# learning rate: fraction of gradient to subtract
lr <- 0.5

for (i in 1:10) {
  
  cat("Iteration: ", i, "\n")
  
  value <- f(param)
  cat("Value is: ", as.numeric(value), "\n")
  
  # compute gradient of value w.r.t. parameter
  value$backward()
  cat("Gradient is: ", as.matrix(param$grad), "\n")
  
  # update
  with_no_grad({
    param$sub_(lr * param$grad)
    # zero out on every iteration (would accumulate otherwise)
    param$grad$zero_()
  })
  
  cat("After update: Param is: ", as.matrix(param), "\n\n")
  
  if (abs(-7 - as.numeric(value)) < 0.00005) break
}

```
